---
title: "Clustering demo"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(palmerpenguins)
```

Pre-process the penguin dataset

```{r}
penguins_dropped <- penguins |> drop_na(bill_length_mm:body_mass_g)
penguins_num <- penguins_dropped |> select(bill_length_mm:body_mass_g) 
```

Check with principal component analysis to decide how many clusters we'd try:

```{r}
penguins_pca <- prcomp(penguins_num, scale.=TRUE)
```


```{r}
ggplot(penguins_pca$x) + geom_point(aes(x=PC1, y=PC2))
```

```{r}
ggplot(penguins_pca$x) + geom_point(aes(x=PC3, y=PC4))
```

Attempt k-means clustering with 2 clusters

```{r}
penguins_km2 <- kmeans(penguins_num, 2)
print(penguins_km2)
```

Construst dataframe with the cluster label included

```{r}
penguins_trans2 <- as.data.frame(penguins_pca$x) |>
  mutate(label=as.factor(penguins_km2$cluster))
```

Visualize the results of clustering:

```{r}
ggplot(penguins_trans2) + 
  geom_point(aes(x=PC1, y=PC2, color=label, shape=label))
```

```{r}
ggplot(penguins_trans2) + 
  geom_point(aes(x=PC3, y=PC4, color=label, shape=label))
```

Retry with 3 clusters (hopefully corresponding to the 3 species)

```{r}
penguins_km3 <- kmeans(penguins_num, 3)
print(penguins_km3)
```

```{r}
penguins_trans3 <- as.data.frame(penguins_pca$x) |>
  mutate(
    label=as.factor(penguins_km3$cluster),
    species=as.factor(penguins_dropped$species)
  )
```

Visualize the results of clustering:

```{r}
ggplot(penguins_trans3) + 
  geom_point(aes(x=PC1, y=PC2, color=label, shape=label))
```

Compare with true labels (species):

```{r}
ggplot(penguins_trans3) +
  geom_point(aes(x=PC1, y=PC2, color=species, shape=species))
```

