---
title: "Principal Component Analysis demo"
author: "Wing-Ho Ko"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggbiplot)
```

Load the Swiss banknotes dataset

```{r}
banknotes <- read_csv("data/banknotes.csv")
```

```{r}
banknotes
```


Perform principal component analysis via `prcomp()`

```{r}
bnotes_pca <- prcomp(
  ~ Length + Left + Right + Bottom + Top, banknotes,
  scale.=TRUE
)
```

Summary information about the PCA

```{r}
print(bnotes_pca)
```

Extract the mean that is being subtracted:

```{r}
bnotes_pca$center
```

Extract the scale factor used (= standard deviation of each variable):

```{r}
bnotes_pca$scale
```

Extract the standard deviation associated with each principal component

```{r}
bnotes_pca$sdev
```

Note that `sum(($sdev)^2) = 5` in our case

```{r}
sum(bnotes_pca$sdev^2)
```

Extract the coordinate transformation matrix programmatically

```{r}
bnotes_pca$rotation
```

Extract the transformed coordinates programmatically

```{r}
as.data.frame(bnotes_pca$x)
```

Scree plot:

```{r}
ggscreeplot(bnotes_pca)
```

Form (unscaled data points and variable vectors) biplot:

```{r}
ggbiplot(bnotes_pca, scale=0, groups=as.factor(banknotes$Currency))
```

Covariance (scaled data points and variable vectors) biplot:

```{r}
ggbiplot(bnotes_pca, scale=1, groups=as.factor(banknotes$Currency))
```

Biplot of the 3rd and 4th principal components:

```{r}
ggbiplot(bnotes_pca, scale=0, choices=c(3, 4), groups=as.factor(banknotes$Currency))
```

