---
title: "UMAP demo"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggbiplot)
library(umap)
```

### Example 1: Swiss banknotes dataset

```{r}
banknotes <- read_csv("data/banknotes.csv")
```

```{r}
banknotes
```

```{r}
bnotes_umap <- umap(
  select(banknotes, Length:Top),
  n_neighbors=15,
  n_components=2,
  min_dist=0.1,
  metric="euclidean"
)
```

Obtain the transformed data points:

```{r}
as.data.frame(bnotes_umap$layout)
```

Put the transformed data points together with currency label into as dataframe

```{r}
bnotes_trans <- as.data.frame(bnotes_umap$layout) |>
  mutate(currency = as.factor(banknotes$Currency))
```

Plot the transformed points

```{r}
ggplot(bnotes_trans, aes(x=V1, y=V2, shape=currency, color=currency)) + 
  geom_point()
```

### Example 2: Trefoil data

Load the simulated trefoil-shaped data

```{r}
trefoil_knot <- read_csv("data/trefoil_knot.csv")
```

Inspect the loaded data

```{r}
trefoil_knot
```

Plot the "top view" (x versus y):

```{r}
ggplot(trefoil_knot, aes(x=x, y=y, color=t)) + geom_point() + 
  coord_fixed() + scale_color_viridis_c(option="turbo")
```
With the `rgl` package we can make an interactive 3d plot of the data 

```{r}
library(rgl)
plot3d(trefoil_knot)
```

Attempt at lower-dimensional representation using PCA:

```{r}
knot_pca <- prcomp(~ x + y + z, trefoil_knot, scale.=FALSE)
ggbiplot(knot_pca, 1:2, scale=0, group=trefoil_knot$t) + 
  scale_color_viridis_c(option="turbo")
```

Try using umap instead:

```{r}
knot_umap <- umap(select(trefoil_knot, x:z), n_neighbors=30)
knot_trans <- as.data.frame(knot_umap$layout) |> mutate(t = trefoil_knot$t)
```

```{r}
ggplot(knot_trans, aes(x=V1, y=V2, color=t)) + geom_point() + 
  scale_color_viridis_c(option="turbo")
```

