---
title: "week 5 exercises"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggbiplot)
library(umap)
library(palmerpenguins)
```

### Exercise 1: covariance and correlation in penguins dataset

Consider the 4 numeric variables in the penguins dataset: bill length, bill depth, flipper length, and body mass. 

```{r}
penguins_sub <- penguins |> select(bill_length_mm:body_mass_g) |> drop_na()
```

1. Compute the covariance matrix between these variables, after dropping the observations with missing measurements

2. Which pair of variables have the most positive covariance? Most negative?

3. Repeat (1) but with correlation matrix.

4. Repeat (2) but with correlations.

### Exercise 2: principal component analysis on limestone

Consider the Nashville carbonates geochemistry dataset, a set of geochemical measurements on Upper Ordovician limestone beds from central Tennessee

```{r}
carbonates <- read_csv("data/nashville_carbonates.csv")
```

Inspect the columns:

```{r}
carbonates
```

We’ll treat the first 2 columns (`strata` and `stratPosition`) as metadata and use the remaining columns as variables. The first two of these (`d13C` and `d18O`) are isotope ratios while the rest are log10 of the weight percentages of major oxides

1. Perform principal component analysis on the dataset, and store the result in `carb_pca`. Make sure you mean-subtracted and rescaled

2. Which variable has the strongest association with the first principal component? How about weakest association?

3. Make a cumulative scree plot (see the `type` argument of `ggscreeplot`). If we want to keep enough principal components so that 90% of the correlation is explained, how many principal components do we need?

4. It turns out that there is a major geological boundary at stratPosition = 34.2. Samples below (smaller stratPosition) this belong to the Carters Formation and samples above (larger stratPosition) are in the Hermitage Formation. 

Make a biplot to visualize the results

5. Based on the plot, Which principal component is useful in distinguishing between rocks from different formations?

### Exercise 3: UMAP on limestone

Consider again the Nashville carbonates geochemistry dataset. Treat the same columns (`d13C` to `log_Si`) as variables and Perform UMAP on the data as follows:

1. Perform UMAP using the default setting. Visualize the result.

2. Try again with `n_neighbors` = 3 and contrast it with the default behavior (`n_neighbors` = 15)

3. Try again with `n_components` = 3 followed by a pca (with `scale.`=FALSE) to extract a 2 dimensional representation. Again contrast it with the default (`n_components` = 2)

### Exercise 4: classifying seeds of wheat

The `seeds_unlabeled.csv` file contains measurements of geometrical properties of kernels belonging to three different varieties of wheat. (Charytanowicz, et. al., DOI:10.1007/978-3-642-13105-9_2). 

1. Perform k-means clustering on the seeds dataset assuming there are 3 groups

2. To visualize the results, perform PCA (with rescaling) on the 7 variables. Then, make a scatter plot with PC1 and PC2 as axes, and color and shape the data point according to the clustering result

3. The actual variety of wheat can be found in the file `seeds_labels.csv`. Make an analogous plot as above where color and shape is determined by the actual wheat type.

4. Find the proportion of seeds that k-means clustering misclassified
